---
title: "Class 11: Protein Structure Prediction with Alphafold2"
author: "Shazreh Hassan (PID: A13743949)"
format: pdf
toc: true
---

## Background

We saw last day that the PDB has 209,886 entries (Oct/Nov 2025). UniProtKB (i.e. protein
sequence database) has 199,579,901 entries.

```{r}
209886/199579901*100
```

The PDB has only 0.1% coverage of the main sequence database.
Enter AlphaFold database (AFDB). < https://alphafold.ebi.ac.uk/ > that attempts to provide
computed models for all sequences in UniProt.
“AlphaFold DB provides open access to over 200 million protein structure predictions to accel-
erate scientific research.”


## Alphafold

AlphaFold has 3 main outputs:

1) The predicted coordinates (PDB files)
2) A local quality score plDDT (one for each amino-acid)
3) A second quality score Predicted Aligned Error PAE (for each pair of amino-acids)

We can run AlphaFold ourselves if we are not happy with AFDB (could be no coverage or poor model)



## Interpreting/analyzing alphafold results in R

```{r}
results_dir <- "HIV_pr_94b5b/"
```

```{r}
pdb_files <- list.files(path=results_dir,
pattern="*.pdb",
full.names = TRUE)

basename(pdb_files)
```

```{r}
library(bio3d)

pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
## library(bio3dview)
## view.pdbs(pdbs)
```

## Alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                       full.names=TRUE)

aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper=TRUE)
```

```{r}
sim <- conserv(aln)
```















