---
title: "Class05: data viz with ggplot"
author: "Shazreh Hassan (PID: A13743949)"
format: pdf
---


Today we are playing with plotting and graphics in R.

There are lots of ways to make figures in R.

**"base" R graphics** (`plot()`, `hist()`), `boxplot()`, etc.)

Add on packages include **ggplot**

```{r}
head(cars)
```

Let's plot this with "base" R:

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot mpg vs disp:

```{r}
#pch of 16 gives filled circles
plot(mtcars$mpg, mtcars$disp, pch=16)
```

Histogram of mpg:

```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function in the ggplot2 is **ggplot()**. The **ggplot2** package needs to be installed. Any package can be installed with the function `install.packages()`. 

> Do NOT run this in the quarto document, run it in console!

```{r}
library(ggplot2)

ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()
```

```{r}
ggplot(cars)+
  aes(speed)+
  geom_histogram()
```

Every ggplot needs at least 3 things:

- the **data** (given with `ggplot(cars)`)
- the **aesthetic** mapping (given with `aes()`)
- the **geom** (given by `geom_point()`)

> For simple canned graphs, "base" R is almost always faster

### Adding more layers

Let's add a line and a title, subtitle, and caption as well as axis labels:
```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()+ 
  geom_smooth(method="lm", se=FALSE)+
  labs(title = "Speed vs distance", 
       x = "Speed (mph)", 
       y = "Distance (ft)",
       caption = "when is tea time anyway??")+
  theme_bw()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

>Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset

>Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```

There are `r sum(genes$State == "up")` UP genes

```{r}
table(genes$State)
```

Make a plot of Condition 1 vs Condition 2, with different states being different colors:

Attempt 1:
```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=Condition2)+
  geom_point()
```

Attempt 2:
```{r}
p <- ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()+
  scale_color_manual(values=c("purple", "gray", "blue"))

p
```

Now that you saved the plot, you can modify it. Add a title:

```{r}
p + labs(title="look at me")
```

Silly example of adding labels:

```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State, label=Gene)+
  geom_point()+
  scale_color_manual(values=c("purple", "gray", "blue"))+
  geom_text()
```

Use ggrepel to put spaces between labels:

```{r}
library(ggrepel)
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State, label=Gene)+
  geom_point()+
  scale_color_manual(values=c("purple", "gray", "blue"))+
  geom_text_repel(max.overlaps = 60)+
  theme_bw()

aes()
```

## Combining Plots

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

```{r}
tail(gapminder)
```

Plot GDP per capita vs life expectancy:
```{r}
#alpha changes transparency - might want to change when there's a lot of overlap in points
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp, col=continent)+
  geom_point(alpha=0.2)+
  facet_wrap(~continent)+
  theme_bw()
```

Changing point size to scale with population size:

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp, col=continent, size=pop)+
  geom_point(alpha=0.2)+
  facet_wrap(~continent)+
  theme_bw()
```

## Bar Chart Practice

Filter for the top 5 countries by population in 2007:
```{r}
library(dplyr)

gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```
Bar chart showing population for these countries:

```{r}
ggplot(gapminder_top5)+
  aes(country, pop)+
  geom_col()
```

